---
title: "Stats 1: An Introduction"
author: "Sara Stoudt (@sastoudt)"
date: "3/8/2020"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading packages

Find out more about the tidyverse [here](https://www.tidyverse.org/).

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
```

## Loading data

Using data from [previous version of this course](https://ireapps.github.io/nicar-2020-schedule#20200306_stats_1_an_introduction_1094_all)

```{r, warning=FALSE, message=FALSE}
boston <- read_csv("boston.csv")
```

Other ways to [load data](https://readr.tidyverse.org/):

- [Excel spread sheets](https://readxl.tidyverse.org/)
- [Google sheets](https://cran.r-project.org/web/packages/googlesheets/googlesheets.pdf)
- Fixed width: `read_fwf()`
- Delimited files: `read_delim()`

## Look at data

```{r, warning=FALSE, message=FALSE}
#View(boston) ## this is a comment
names(boston)
dim(boston)
head(boston)
str(boston)
```

## Summaries of Continuous Variables

```{r, warning=FALSE, message=FALSE}
## continuous
boston %>% select(mph) %>% summary()
boston %>% select(mphover) %>% summary()
```

## Summaries of Discrete Variables

```{r, warning=FALSE, message=FALSE}
## discrete
boston %>% select(sex) %>% table()
boston %>% select(race) %>% table()
boston %>% select(daynight) %>% table()
boston %>% select(ticket) %>% table()
```

## Two-Way Summaries

```{r, warning=FALSE, message=FALSE}
## two-table
boston %>% select(sex, race) %>% table()
boston %>% select(sex, ticket) %>% table()
boston %>% select(race, ticket) %>% table() 
```

## Summaries by Group

Find out more about summaries [here](https://dplyr.tidyverse.org/).

```{r, warning=FALSE, message=FALSE}
boston %>%
  group_by(sex) %>%
  summarise(meanSpeeding = mean(mphover))
boston %>%
  group_by(daynight) %>%
  summarise(meanSpeeding = mean(mphover))
boston %>%
  group_by(ticket) %>%
  summarise(meanSpeeding = mean(mphover), minSpeeding = min(mphover), maxSpeeding = max(mphover))

## two way
boston %>% group_by(ticket, zone) %>% summarise(count = n()) %>% arrange(zone)
```

## Transforming Variables

**[Factors](https://forcats.tidyverse.org/)**

```{r, warning=FALSE, message=FALSE}
## make ticket factor
class(boston$ticket)
boston$ticket <- as.factor(boston$ticket)
class(boston$ticket)
```

```{r}
## group race --> minority

boston <- boston %>% mutate(minority = case_when(
  race == "W" ~ 0,
  race != "W" ~ 1
))

## recode sex --> female

boston <- boston %>% mutate(female = case_when(
  sex == "F" ~ 1,
  sex == "M" ~ 0
)) ## U will become NA

```

## [Plots](https://ggplot2.tidyverse.org/)

```{r, warning=FALSE, message=FALSE}
## mph over by ticket
ggplot(boston, aes(x = mphover, color = ticket, fill = ticket)) +
  geom_histogram(alpha = 0.6, position = "identity")

## three variables
ggplot(boston, aes(x = mphover, color = ticket, fill = ticket)) +
  geom_histogram(alpha = 0.6, position = "identity") + facet_wrap(~female)
```

## Statistical Tests

**Is the average miles per hour over the speed limit different for ticket and non-tickets?**

```{r}
t.test(mphover ~ ticket, data = boston)
```

[Assumptions](https://en.wikipedia.org/wiki/Student%27s_t-test#Assumptions)

TLDR

- large sample size (at least >30)
- independent samples (independence between drivers)

**Is the probability of getting a ticket for a female different than the probability of getting a ticket for a non-female?**

```{r}
prop.test(table(boston$ticket, boston$female), correct=FALSE) 
```

Assumptions

- observations are random sample of population
- independent observations
- have at least 10 successes and 10 failures.

*Healthy skepticism*: What if females were more likely to drive during the day and tickets are more likely to be given out at night?

To deal with this kind of thing stick around for [Linear Regression](https://ireapps.github.io/nicar-2020-schedule#20200308_stats_2_linear_regression_repeat_1098_all).
