---
title: "Stats 1: An Introduction"
author: "Sara Stoudt (@sastoudt)"
date: "3/8/2020"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading packages

Find out more about the tidyverse [here](https://www.tidyverse.org/).

```{r, warning=FALSE, message=FALSE}
#install.packages("tidyverse")
library(tidyverse)
```

## Loading data

Using data from [previous version of this course](https://ireapps.github.io/nicar-2020-schedule#20200306_stats_1_an_introduction_1094_all)

```{r, warning=FALSE, message=FALSE}
boston <- read_csv("boston.csv")
```

Other ways to [load data](https://readr.tidyverse.org/):

- [Excel spread sheets](https://readxl.tidyverse.org/)
- [Google sheets](https://cran.r-project.org/web/packages/googlesheets/googlesheets.pdf)
- Fixed width: `read_fwf()`
- Delimited files: `read_delim()`

## Look at data

```{r, warning=FALSE, message=FALSE}
names(boston)
dim(boston)
head(boston)
str(boston)
#View(boston) ## this is a comment
```

## Summaries of Continuous Variables

```{r, warning=FALSE, message=FALSE}
## continuous
boston %>% select(mph) %>% summary()
boston %>% select(mphover) %>% summary()
```

## Summaries of Discrete Variables

```{r, warning=FALSE, message=FALSE}
## discrete
boston %>% select(sex) %>% table()
boston %>% select(race) %>% table()
boston %>% select(daynight) %>% table()
boston %>% select(ticket) %>% table()
```

## Two-Way Summaries

```{r, warning=FALSE, message=FALSE}
## two-table
boston %>% select(sex, race) %>% table()
boston %>% select(sex, ticket) %>% table()
boston %>% select(race, ticket) %>% table() 
```

## Summaries by Group

Find out more about summaries [here](https://dplyr.tidyverse.org/).

```{r, warning=FALSE, message=FALSE}
boston %>%
  group_by(sex) %>%
  summarise(meanSpeeding = mean(mphover))
boston %>%
  group_by(daynight) %>%
  summarise(meanSpeeding = mean(mphover))
boston %>%
  group_by(ticket) %>%
  summarise(meanSpeeding = mean(mphover), minSpeeding = min(mphover), maxSpeeding = max(mphover))

## two way
boston %>% group_by(ticket, zone) %>% summarise(count = n()) %>% arrange(zone)
```

## Transforming Variables

**[Factors](https://forcats.tidyverse.org/)**

```{r, warning=FALSE, message=FALSE}
## make ticket factor
class(boston$ticket)
boston$ticket <- as.factor(boston$ticket)
class(boston$ticket)
```

```{r}
## group race --> minority

boston <- boston %>% mutate(minority = case_when(
  race == "W" ~ 0,
  race != "W" ~ 1
))

boston %>% select(minority) %>% table()

## recode sex --> female

boston <- boston %>% mutate(female = case_when(
  sex == "F" ~ 1,
  sex == "M" ~ 0
)) ## U will become NA

boston %>% select(female) %>% table()

boston <- boston %>% mutate(mphover = mph - zone)

boston %>% select(mphover) %>% summary()
```

## [Plots](https://ggplot2.tidyverse.org/)

```{r, warning=FALSE, message=FALSE}

ggplot(boston, aes(x = mphover)) +
  geom_histogram(position = "identity")

## mph over by ticket
ggplot(boston, aes(x = mphover, color = ticket, fill = ticket)) +
  geom_histogram(alpha = 0.6, position = "identity")

## three variables
ggplot(boston, aes(x = mphover, color = ticket, fill = ticket)) +
  geom_histogram(alpha = 0.6, position = "identity") + facet_wrap(~female)
```

## Statistical Tests

**Is the average miles per hour over the speed limit different for ticket and non-tickets?**

*difference in means test*

```{r}
t.test(mphover ~ ticket, data = boston)
```

*Assumptions*

- large sample size (at least >30)
- independent samples (independence between drivers)

*Null hypothesis:* There is no significant difference in the average miles per hour over the speed limit for tickets and non-tickets.

*Alternative hypothesis:* There is a significant difference in the average miles per hour over the speed limit for tickets and non-tickets.

*Interpretation:* If the average miles per hour over the speed limit was the same for tickets and non-tickets, we would get a difference in means equal to or more extreme to the one we see here with probablility 0.0005 (p-value). This is unlikely, providing evidence for the hypothesis that there is a difference.

*Healthy skepticism*: Is the difference practically significant?

**Is the proportion of females getting tickets different than the proportion of males getting tickets?**

*difference in proportions test*

```{r}
prop.test(table(boston$ticket, boston$female)) 
```

*Assumptions*

- observations are random sample of population
- independent observations
- have at least 10 successes (ticket) and 10 failures (no ticket)

*Null hypothesis:* There is no significant difference in proportion of males and females who got tickets.

*Alternative hypothesis:* There is a significant difference in proportion of males and females who got tickets.

*Interpretation:* If the proportion of males and females who got tickets were the same, we would get a difference in proportions equal to or more extreme to the one we see here with probablility 0.068 (p-value). This is bigger than a typical cutoff used (0.05), providing less evidence for the hypothesis that there is a difference.

*Healthy skepticism*: What if females were more likely to drive during the day and tickets are more likely to be given out at night? Also what is so magical about the 0.05 p-value cut-off?

To deal with this kind of thing stick around for [Linear Regression](https://ireapps.github.io/nicar-2020-schedule#20200308_stats_2_linear_regression_repeat_1098_all).
